Aligning species range data to better serve science and conservation

### 2016-01-13 data_explore/data_explore.Rmd
updating manuscript, and incorporating info on AM occurrence cells

### mongo plot time
scatter_arearatio_vs_overlap <- ggplot(spp_list %>% 
                                         transform(popn_category = reorder(popn_category, category_score)),
                                       aes(x = area_ratio, y = sm_perc, 
                                           color = popn_category,
                                           size = lg_area_pct)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "green4")  + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "greenyellow") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orange") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orangered2") + 
  geom_point(alpha = .6)

>>>>

spp_list <- spp_list %>% mutate(data_poor = occurcells < 10)
### mongo plot time
scatter_arearatio_vs_overlap <- ggplot(spp_list %>% 
                                         transform(popn_category = reorder(popn_category, category_score)),
                                       aes(x = area_ratio, y = sm_perc, 
                                           color = data_poor,
                                           size = lg_area_pct)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "green4")  + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "greenyellow") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orange") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orangered2") + 
  geom_point(alpha = .6)

### 2016-01-13 data_explore/data_prep.Rmd
updating manuscript, and incorporating info on AM occurrence cells

spp_list <- spp_list %>%
  filter(!is.na(sm_perc)) %>%
  ### will ditch the ones where no IUCN polygon exists - or are these species
  ### whose ranges only occur in cells beyond the LOICZID grid?
  filter(!str_detect(spatial_source, 'subpop') & !str_detect(spatial_source, 'alias'))
spp_list <- spp_list %>%
  left_join(read.csv(file.path(dir_data, 'sciname_lookups/spp_groupnames.csv'), 
                     stringsAsFactors = FALSE),
            by = 'spp_group')

write.csv(spp_list, file.path(dir_data, 'spp_list_w_area_trimmed.csv'))

>>>>

spp_list <- spp_list %>%
  filter(!is.na(sm_perc)) %>%
  ### will ditch the ones where no IUCN polygon exists - or are these species
  ### whose ranges only occur in cells beyond the LOICZID grid?
  filter(!str_detect(spatial_source, 'subpop') & !str_detect(spatial_source, 'alias')) 
spp_list <- spp_list %>%
  left_join(read.csv(file.path(dir_data, 'sciname_lookups/spp_groupnames.csv'), 
                     stringsAsFactors = FALSE),
            by = 'spp_group') %>%
  left_join(spp_am %>% select(am_sid = speciesid, occurcells),
            by = 'am_sid')

write.csv(spp_list, file.path(dir_data, 'spp_list_w_area_trimmed.csv'), 
          row.names = FALSE)

### 2016-01-13 ms_1_abst_sig.Rmd
updating manuscript, and incorporating info on AM occurrence cells

## Abstract

_get to this later..._

>>>>

## Abstract
<!--
Mapping and predicting species distributions is fundamental to the sciences of ecology, biogeography, and conservation, among many others.  Knowing where individuals of a species exist, and where they thrive, provides foundational information for 
 understanding species ranges and diversity,
 predicting species responses to human impacts and climate change,
 and managing and protecting species effectively.
A rich literature tackles the many dimensions of these questions.
-->
- Why are species distribution maps important

<!--
Two important repositories _predict_ marine species ranges throughout the world's oceans – AquaMaps modeled species distribution maps (Kaschner et al. 2013) and International Union for Conservation of Nature (IUCN) species range maps (REF).  These spatial datasets are used for a wide range of purposes, including assessing marine species status (Halpern et al. 2012, Selig et al. 2013), evaluating global biodiversity patterns (Coll et al. 2010, Martin et al. 2014), predicting range shifts (Molinos et al. 2015), and setting conservation priorities (Klein et al. 2015). 
-->
Species maps from AquaMaps and the International Union for Conservation of Nature describe marine species ranges throughout the world's oceans, and are widely used in a wide range of biodiversity and conservation purposes. ___get across the point that these two datasets are the big ones___

<!--
The two datasets ostensibly describe the same information, but significant differences in methodology and intent could lead to dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations.

Importantly, biases in taxonomic or spatial coverage within a dataset could shift management or conservation actions away from places or species that are most in need.(maybe Jetz 2008?)  Inaccurate indications of presence or absence could lead to inefficient or ineffective marine reserve systems and management plans. (Rondinini et al. 2006) (Jetz 2008?)
-->

- These data sets are different, due to intent and methodology.
- Spatial and taxa coverages are important to understand impacts on different analyses 
    - taxa (probs) drives spatial
    - small overlap provides opportunity to increase species without much redundancy
- Comparisons of maps for overlapping species informs possible improvements
    - improve method for outlining IUCN maps
    - identify common problems and point toward solutions
- How differences can affect outcomes of research and conservation studies

<!--
To understand the implications of differences between these two datasets, we compared how each data set represents the global spatial and taxonomic distribution of species.  For the relatively small number of species mapped in both datasets, we examined how well the species maps align. We then subjected two recent marine biodiversity studies - the Ocean Health Index biodiversity goal (Halpern et al. 2012) and a global MPA gap analysis (Klein et al. 2015) - to a sensitivity analysis, substituting one dataset over the other, to highlight the possible consequences of different data use decisions on our understanding of _the status of_ marine biodiversity.
-->

### 2016-01-13 ms_2_intro_res_disc.Rmd
updating manuscript, and incorporating info on AM occurrence cells

___Peer? Dive? Cast a scientific eye?___ _beneath the waves anywhere in the world - a moody kelp forest, a bustling coral reefscape, the frigid depths beneath a polar ice sheet, the endless blue of the open ocean.  Which species would you expect to find, and which species would be woefully out of place?_ Mapping and predicting species distributions is fundamental to the sciences of ecology, biogeography, and conservation, among many others.  Knowing where individuals of a species exist, and where they thrive, provides foundational information for 
 understanding species ranges and diversity,
 predicting species responses to human impacts and climate change,
 and managing and protecting species effectively.
A rich literature tackles the many dimensions of these questions.

One major outcome of this body of science is the various compiled databases of species distribution maps. <!--In the oceans there are now two such global repositories--> Two important repositories _predict_ marine species ranges throughout the world's oceans  <!--_(if we define it narrowly; OBIS is global species occurrences, but not quite range or distribution maps)_--> – AquaMaps modeled species distribution maps (Kaschner et al. 2013) and International Union for Conservation of Nature (IUCN) species range maps (REF).  These spatial datasets are used for a wide range of purposes, including assessing marine species status (Halpern et al. 2012, Selig et al. 2013), evaluating global biodiversity patterns (Coll et al. 2010, Martin et al. 2014), predicting range shifts (Molinos et al. 2015), and setting conservation priorities (Klein et al. 2015). The two datasets ostensibly describe the same information, but significant differences in methodology and intent could lead to dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations.

Importantly, biases in taxonomic or spatial coverage within a dataset could shift management or conservation actions away from places or species that are most in need.(maybe Jetz 2008?)  Errors of commission (false indications of presence) or omission (false indications of absence) could lead to inefficient or ineffective marine reserve systems and management plans. (Rondinini et al. 2006) (Jetz 2008?)

@@

IUCN range maps are intended to include all possible regions in which a species may be present, with the caveat that this does not imply the species is evenly distributed everywhere within the boundaries of the map(REF).  AquaMaps modeled distribution maps provide a more nuanced prediction of species presence within the extent of occurrence, but the model may not capture all the complexities that drive species distribution (REF).  Due to these differences in methodology and intent, IUCN range maps are more likely to overpredict species presence while AquaMaps modeled distribution maps are more likely to underpredict species presence.(REF and REF)

@@

Quadrant 1 contains species whose map pairs agree in both spatial distribution and the extent of described ranges, as we would expect for well-understood species.  Excellent and valuable, but not particularly interesting.  The extent-misaligned maps contained in quadrant 2, however, provide a richer set of examples to understand the fundamental differences between the datasets; examining these, we can identify several mechanisms that could cause such extent misalignment.

For `r n_q2_am_sm` of `r n_q2` species in this quadrant (`r round((n_q2_am_sm/n_q2) * 100, 1)`%), the IUCN extent is larger than the AquaMaps extent.  By itself, this is not surprising; but for many of these species, a quick look at the maps (see SOM for examples) shows that the IUCN range hews closely to the AquaMaps range, while adding a significant buffer zone.  Coral species dominate this quadrant (`r n_q2_corals` coral species - `r round(n_q2_corals/n_corals * 100, 1)`% of all corals and `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in quadrant 2).  We suspect that many of these extent-misaligned map pairs can be easily explained: most corals and reef-associated organisms prefer shallower waters; seafloor depth is explicitly modeled in AquaMaps, but not explicitly considered in IUCN range considerations. ___include recommendation for IUCN range maps to be clipped to bathymetry where appropriate? or save til end?___

___wrap up something with quadrants 3 and 4:___ _In several instances, the AquaMaps map conformed to a small, local subset of the IUCN map (see SOM for examples - what was the Q4 graph?), in some cases due to some point locality observations being rejected by AquaMaps experts (EXAMPLES) and in other cases due to differences in species identification._ ___bleh - this needs help. Evidence? examples?___

Poorly-aligned rangemaps, regardless of taxon or extinction risk category, indicate species that could benefit from further expert study. ___is this just 'duh'? is there something more important I can say?___

@@

* Users
    * understand that range maps tend to overestimate presence; AquaMaps are less likely to overestimate presence, but trade off for overestimating absence. Both better than simple point locality data.
    * to use both together:
        * consider setting AM presence threshold to 0% to create an extent of occurrence map, more like IUCN map.
        * consider de-weighting IUCN for area-weighted calcs, for taxa in which IUCN is consistently overestimating presence

>>>>

_Peer beneath the waves anywhere in the world - a moody kelp forest, a bustling coral reefscape, the frigid depths beneath a polar ice sheet, the endless blue of the open ocean.  Which species would you expect to encounter, and which species would be woefully out of place?_ Mapping and predicting species distributions is fundamental to the sciences of ecology, biogeography, and conservation, among many others.  Knowing where individuals of a species exist, and where they thrive, provides foundational information for 
 understanding species ranges and diversity,
 predicting species responses to human impacts and climate change,
 and managing and protecting species effectively.
A rich literature tackles the many dimensions of these questions.

One major outcome of this body of science is the various compiled databases of species distribution maps. <!--In the oceans there are now two such global repositories--> Two important repositories _predict_ marine species ranges throughout the world's oceans – AquaMaps modeled species distribution maps (Kaschner et al. 2013) and International Union for Conservation of Nature (IUCN) species range maps (REF).  These spatial datasets are used for a wide range of purposes, including assessing marine species status (Halpern et al. 2012, Selig et al. 2013), evaluating global biodiversity patterns (Coll et al. 2010, Martin et al. 2014), predicting range shifts (Molinos et al. 2015), and setting conservation priorities (Klein et al. 2015).

The two datasets ostensibly describe the same information, but significant differences in methodology and intent could lead to dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations.

Importantly, biases in taxonomic or spatial coverage within a dataset could shift management or conservation actions away from places or species that are most in need.(maybe Jetz 2008?)  Inaccurate indications of presence or absence could lead to inefficient or ineffective marine reserve systems and management plans. (Rondinini et al. 2006) (Jetz 2008?)

@@

IUCN range maps are intended to include all possible regions in which a species may be present, with the caveat that this does not imply the species is evenly distributed everywhere within the boundaries of the map(REF).  AquaMaps modeled distribution maps provide a more nuanced prediction of species presence within the extent of occurrence, but the model may not capture all the complexities that drive species distribution (REF).  Due to these differences in methodology and intent, IUCN range maps are more likely to overpredict species presence while AquaMaps modeled distribution maps are relatively more likely to underpredict species presence.(REF and REF)

@@

Quadrant 1 contains species whose map pairs agree in both spatial distribution and the extent of described ranges, as we would expect for well-understood species.  Excellent and valuable, but not particularly interesting.

The extent-misaligned maps contained in quadrant 2 provide a richer set of examples to understand the fundamental differences between the datasets; examining these, we can identify several mechanisms that could cause such extent misalignment.

For `r n_q2_am_sm` of `r n_q2` species in this quadrant (`r round((n_q2_am_sm/n_q2) * 100, 1)`%), the IUCN extent is larger than the AquaMaps extent.  By itself, this is not surprising; but for many of these species, a quick look at the maps (see SOM for examples) shows that the IUCN range hews closely to the AquaMaps range, while adding a significant buffer zone.  Coral species dominate this quadrant (`r n_q2_corals` coral species - `r round(n_q2_corals/n_corals * 100, 1)`% of all corals and `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in quadrant 2).  We suspect that many of these extent-misaligned map pairs can be easily explained: most corals and reef-associated organisms prefer shallower waters; seafloor depth is explicitly modeled in AquaMaps, but not explicitly considered in IUCN range considerations. (For examples and details, see SOM) ___include recommendation for IUCN range maps to be clipped to bathymetry where appropriate? or save til end?___

The extent-alignment of species maps found in quadrant 3 can in many cases be attributed to "two wrongs make a right."  For these species, IUCN ranges are frequently overextended into unsuitable habitat, as in the case of many quadrant 2 species, but at the same time, AquaMaps aggressively extrapolates distribution of presence into locations where IUCN predicts absence.  This results in an area ratio close to 1, giving the impression of extent alignment. ___wrap up something with quadrants 3 and 4:___ _In several instances, the AquaMaps map conformed to a small, local subset of the IUCN map (see SOM for examples - what was the Q4 graph?), in some cases due to some point locality observations being rejected by AquaMaps experts (EXAMPLES) and in other cases due to differences in species identification._ ___bleh - this needs help. Evidence? examples?___

Quadrant 4 species maps seem to suffer from the same problem as those in quadrant 3, but here the AquaMaps extrapolation extends even further, pushing the extent alignment ratio lower.  In this quadrant, AquaMaps predicts a larger extent than IUCN for 70% of species represented in this quadrant, compared with just 30% of species in the entire set of overlapping maps.

_how many species are "reviewed" in each quadrant? how many are data poor?_
overall: 21% of overlapping species are data poor (model based on fewer than 10 cell-occurrences)
q1: 8.2% (33/401); q2: 22.1% (151/682); q3: 16.5% (114/692); q4: 39.7% (159/401)
Overall: 13.4% of species reviewed...
q1: 24.9%; q2: 14.7%; q3: 9.5%; q4: 6.2%

@@

* Users
    * understand that range maps tend to overestimate presence; AquaMaps are less likely to overestimate presence, but trade off for overestimating absence. Both better than simple point locality data.
    * to use both together:
        * consider setting AM presence threshold to 0% to create an extent of occurrence map, more like IUCN map.
        * consider additional filters and criteria, e.g. "reviewed" or "occurcells" 
        * consider clipping IUCN ranges to bathymetry for depth-limited species or taxa
        * or consider de-weighting IUCN for area-weighted calcs, for taxa in which IUCN is consistently overestimating presence

### 2016-01-13 ms_6_supplemental.Rmd
updating manuscript, and incorporating info on AM occurrence cells

>>>>

species to look more closely at:
a few codes to indicate likely problems?
- DC = depth clipping; IUCN extents go beyond reasonable depth for the species
- DX = data excluded; AM excludes some observations that appear to be used for IUCN
- DP = data poor (AquaMaps acknowledges data poor status)
- NX = needs evaluation by expert; AM model likely predicts areas where species is unlikely to be found based on observed occurrences - does AM model reflect same as AM suitable habitat?

Q2:
Conus episcopatus (cones) DP, DX - d-match around Coral Triangle, but IUCN shows much more around Indian Ocean and S Pacific; also IUCN extents
Ctenochaetus binotatus - IUCN extents beyond AM
Naso vlamingii (acanthuridae) - IUCN extents beyond AM
Thalassoma purpureum (wrasses) - IUCN extents beyond AM
Porites nigrescens (corals 2) - DP - AM limited to CT; IUCN shows far greater dist, and extents
Conus ammiralis (cones) - DP - AM limited to CT, IUCN shows in Madagascar; also extents
Conus tessulatus (cones) - extents
Acropora sarmentosa (corals 1) - extents; also AM shows HI but IUCN does not?
Holothuria fuscogilva (sea cucumbers) - extents; also AM shows Mex and Central Am west coast, but IUCN does not (not available online?)
Oculina Varicosa (corals 2) - lim to Caribbean; extents - this might be a good close-up map
Acanthocybium solandri (tunas/billfish) - mostly tropical by IUCN; AM shows much farther N and S

Q3:
Conus magnificus (q3 - IUCN distribution is more limited than AM distribution)
Abudefduf concolor (damselfish) - Baja for AM, southern Central Am for IUCN; maybe a good closeup map?  AM points from Baja to Peru incl Galapagos; some rejected pts from Caribbean and eastern Atlantic
Centropyge aurantonotus (angelfish) - AM shows Caribbean; IUCN shows northern South America, also maybe a good closeup map? AM points from FL to south Brazil, but no Gulf of Mexico.
Chlorurus perspicillatus (parrotfish) - IUCN shows only HI; AM shows dots all over S Pacific and near Marianas Islands
Sarda orientalis (tunas/billfish) - limited dists by IUCN
Montastraea franksi (corals 2) - DP - IUCN shows just in Caribbean; AM shows all up & down east coast of N America and S America

Q4:
Acanthurus nigroris (q4 - IUCN dist limited to HI, while AM shows all over S Pacific and CT) - AM observations all over S pacific...
Praealticus tanegasimae (blenny) - IUCN shows limited dist around Japan/Marianas; AM shows much broader through CT and W Pacific - good local map?
Stethojulis marquesensis (wrasse) - IUCN shows only small cluster of islands (Marquesas?) while AM shows much broader across S Pacific
Canthigaster leoparda (puffer) - IUCN limited spots around a couple of islands; AM shows all throughout CT
pentacheles snyderi (lobster) - DP - AM shows all over - sim to depth profiles? while IUCN shows only a couple patches - HI? S Pacific, and S Indian oceans?  AM limits to specific FAO regions, which results in odd cutoffs; also shows some data points in N Atlantic which has an odd result.
Nephropsis sulcata (lobster) - IUCN shows odd ranges, most of which don't overlap with AM
